### 2012-11-29 paper/isca40format/DistributedMsi.tex
before reorging scheduler etc

Using these commands, the designer can now easily specify a distributed MSI
protocol, by specifying the methods \uReq{} and \dReq{}. Three other
obligations to take care of:
\begin{enumerate}
\item Handler \uReq{}($c, a, x$) can be called even if $x \ge
this.dir[c]$. In this case, it should be dropped.
\item Handler \dReq{}($p, a, x$) can be called even if $x \le
this.state[a]$. In this case, it should be dropped.
\item A new handler, \dRespL{}($c, a, x$), where $c$ is a child of cache $this$,
should be provided, and this informs $this$ that $c$ has changed its state to
$x$.
\end{enumerate}

Using the above set of rules, the designer can specify the distributed MSI protocol as shown in
Figure \ref{realistic}. Handlers \uReq{} and \dReq{} are exactly the same as
shown in Figure \ref{msi-template}, except that whenever the state of a child
$c'$ is read, it should be replaced syntactically by $this.dir[c'][a]$.

@@

\begin{algorithmic}
\State // Upgrading $this$ by sending request to parent and getting a response
\Proc {\uReqL}{$p, a, x$}
  \If {$this$ is LLC \&\& $this.state[a] = I$}
    \State \send{} \Req{this}{Memory}{a}{M};
    \State \receive{} \Data{Memory}{this}{a}{d};
    \State $this.data[a] \gets d$;
    \State $this.state[a] \gets M$;
  \Else
  \State \send{} \Req{this}{p}{a}{x};
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndIf
\EndProc
\State // Downgrading $this$' child by sending request to child and getting a
response
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  \EndWhile
\EndProc
\State // $this$ finally upgrading the directory state of its child and sending
a response to its child
\Proc {\uResp}{$c, a, x$}
  \State \textbf{if} ($this.dir[c][a] = I$)
  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
  \State $this.dir[c][a] \gets x$;
\EndProc
\State // $this$ finally downgrading its state and sending a response to its
parent
\Proc {\dResp}{$p, a, x$}
  \State \textbf{if} ($isModified(this.state[a])$)
  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\caption{Distributed MSI protocol. In all the methods $p$ is the parent of
$this$, $c$ is a child of $this$, and the methods work on address $a$ to change
its state (or directory) to $x$. This function only shows the auxiliary
functions and the handler for voluntary responses from children. The handler
for requests are exactly the same as in Figure \ref{msi-template}, with every
occurrence of $c'.state[a]$ replaced by $this.dir[c'][a]$}
\label{realistic}
\end{figure}

@@

Now that we have established the vocabulary of threads, we will proceed to
describe the requirements that a network should guarantee in order to make such
an implementation possible. We will also show the scheduling algorithm that we
generate.

\subsection{Message ordering requirements}
\floatstyle{boxed}
\restylefloat{figure}

@@

\subsection{Resource requirements and blocking restrictions}

\begin{figure}\small
\begin{requirement}
At a cache node, an incoming request from the parent should not be blocked by any
incoming request from its children.\label{cReqNoBlockPReq}
\end{requirement}
\begin{requirement}
At a cache node, an incoming response from any of its children should not
be blocked by any incoming requests.\label{reqNoBlockResp}
\end{requirement}
\begin{requirement}
At a cache node, an incoming response for address $a$ from its parent should not be blocked by
an incoming request from its child for address $a$. \label{reqBlockResp2}
\end{requirement}
\caption{Blocking restrictions}
\label{blocking}
\end{figure}

Figure \ref{blocking} shows the restrictions that the input channels should
obey with respect to an incoming message blocking other incoming messages.

\begin{figure}\small
\begin{requirement}
Each cache node has at least one dedicated thread resource for handling only messages
(requests or responses) from the parent\label{dedicate1}
\end{requirement}
\begin{requirement}
Each cache node has at least one dedicated thread resource for handling only responses
(from children or from the parent)\label{dedicate2}
\end{requirement}
\begin{requirement}
Each cache node has at least three thread resources\label{dedicate3}
\end{requirement}
\caption{Dedicated resources in a cache node}
\label{dedicated}
\end{figure}

Figure \ref{dedicated} shows the minimum number of thread resources that should
be present in each cache node. Lack of required thread resouces leads to the same
kind of deadlock scenarios as incorrect blocking does. This is illustrated in the
following:

Let's say both $c_1.state[a] = p.dir[c_1][a] = S$ and $c_2.state[a] =
p.dir[c_2][a] = S$. Both $c_1$ and $c_2$ get store requests from the processor
for address $a$, and they send upgrade-to-$M$ requests to $p$.  $p$ receives the
request from $c_2$ first. $p$ sends a downgrade-to-$I$ request to $c1$.

Let's say Requirement \ref{cReqNoBlockPReq} is not held, and instead some request
from the core is blocking $c_1$ from receiving the request. $c_1$ will never
respond back to $p$. But $p$ will be waiting in procedure \dReqL{}, and will
never start handling the request from $c_1$. This will create a deadlock.  The
same deadlock is also created when $c_1$ is out of thread resources to create a
thread to handle the downgrade request from $p$ (violating Requirement
\ref{dedicate1}). Figure \ref{unknown} depicts this scenario.

Let's extend this scenario further. Instead of violating Requirements
\ref{cReqNoBlockPReq} or \ref{dedicate1}, let $c_1$
send a downgrade-to-$I$ response to $p$. Let's instead assume that Requirement
\ref{reqNoBlockResp} is violated in $p$. So, the downgrade-to-$I$ response from
$c_1$ is not received by $p$, leading to the same deadlock. This is the same if
Requirement \ref{dedicate2} is violated and $p$ is out of thread resources to
create a thread to handle the downgrade response.

The need for Requirement \ref{dedicate3} can be shown easily. Let's say that a
cache has exactly 2 thread resources, one for handling messages from parents,
and one for handling responses from either the parent or the children. Now, if
it receives a request from one of its children, it can not use either of the
thread resources. Thus, at least one more thread resource is needed for
handling requests from children. This thread resource can also be used for
handling other types of incoming message.

\subsection{Scheduler requirements}

We will now list the requirements that a scheduler must obey to give the
illusion of non-suspensive threads and avoid deadlocks.

\begin{figure}\small
\begin{requirement}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{requirement}
\begin{requirement}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from $p$, or
\item a thread in $c$ is handling a request for address
$a$ from one of $c$'s children $c'$, but that thread is not waiting for a
response from $p$
\end{enumerate}
\label{cHandleReq}
\end{requirement}
\caption{Requirements for receiving a request (used by the scheduler)}\label{recvReq}
\label{}
\end{figure}

\begin{figure}\small
\begin{requirement}
Address $a$ present in cache node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}\label{evict}
\caption{Requirements for suspending a thread during choosing an address for eviction}
\end{requirement}
\end{figure}

Requirements \ref{pHandleReq}, \ref{cHandleReq} and \ref{evict} are needed
because once a thread has finished up-to a point in its execution, it can not
re-execute that portion. For instance, consider a thread handling an upgrade
request from a child. Let's say thread $t$ has progressed to the point where it
is simply waiting to send an upgrade response to the child. Now, if the line
gets evicted because it was chosen for replacement, or it got downgraded by the
parent, then thread $t$ can not re-execute its procedure again. To prevent this,
only one thread is allowed to operate on a particular address, except in the
case of downgrade requests from the parent. In this case, a thread is created to
handle the downgrade request from the parent even if there is another thread
handling a request for the same address from a child, if the latter thread is
just waiting for a response from the parent. This is done to avoid deadlocks
like the ones shown in Figure \ref{unknown}.

\begin{figure}\small
\begin{requirement}
Starvation-freedom: A thread that keeps becoming ready to execute will
eventually be scheduled to execute.\label{starvation}
\end{requirement}
\end{figure}

Local starvation-freedom is required in each cache node. If a cache node never
executes a particular suspended thread, even if it keeps getting ready to
execute, then the thread will not progress, leading to a deadlock.

A scheduler can be implemented as described in Figure \ref{scheduler}, obeying
Requirements \ref{pHandleReq} and \ref{cHandleReq}. It should perform some form of
fair arbiration among its suspended threads to avoid starvation (Requirement
\ref{starvation}).

\floatstyle{plain}
\restylefloat{figure}

\begin{figure}
\small
\begin{boxedminipage}{\linewidth}
\begin{alltt}
\normalfont
\lWhile{} (\(True\)) \bopen
      \lIf (\Resp{c}{n}{a}{x} is in the input channel,
            where \(n = c.parent\)) \bopen
            \lIf (suspended thread \(t\) is waiting for 
                 a response from \(c\) for address \(a\)) \bopen
                   \resume{} \(t\);
            \bclose \lElse \bopen
                   \receive{} \Resp{c}{n}{a}{x};
                   \start{} \dRespL(\(c, n, a, x\));
            \bclose
      \bclose \lElsIf (\Resp{p}{n}{a}{x} is in the input channel,
            where \(p = n.parent\)) \bopen
                   // There must be a suspended thread \(t\)
                   waiting for a response from \(p\) for address \(a\)
                   \resume{} \(t\);
            \bclose
      \bclose \lElsIf (\Req{c}{n}{a}{x} is in the input channel,
                     where \(n = c.parent\)) \bopen
             \lIf (Local-Property \ref{pHandleReq} is satisfied) \bopen
                   \receive{} \Req{c}{n}{a}{x};
                   \start{} \uReq(\(c, n, a, x\));
             \bclose
      \bclose \lElsIf (\Req{p}{n}{a}{x} is in the input channel,
                  where \(p = n.parent\)) \bopen
             \lIf (Local-Property \ref{cHandleReq} is satisfied) \bopen
                   \receive{} \Req{p}{n}{a}{x};
                   \start{} \dReq(\(p, n, a, x\));
             \bclose
      \bclose \lElsIf (suspended thread \(t\) is ready to execute) 
             \resume{} \(t\);
\bclose
\end{alltt}
\end{boxedminipage}
\caption{Implementation of a scheduler for cache node $n$}
\label{scheduler}
\end{figure}

>>>>

Using these commands, the designer can now easily specify a distributed MSI
protocol, by specifying the methods \uReq{} and \dReq{}. Three other
obligations to take care of:
\begin{enumerate}
\item Handler \uReq{}($c, a, x$) can be called even if $x \ge
this.dir[c]$. In this case, it should be dropped.
\item Handler \dReq{}($p, a, x$) can be called even if $x \le
this.state[a]$. In this case, it should be dropped.
\item A new handler, \dRespL{}($c, a, x$), where $c$ is a child of cache $this$,
should be provided, and this informs $this$ that $c$ has changed its state to
$x$. It is the designer's responsibility to update $this.dir[c][a]$ and his
copy of the data if need be.
\end{enumerate}

Using the above set of rules, the designer can specify the distributed MSI protocol as shown in
Figures \ref{msi-unsolicited} and \ref{realistic}. Handlers \uReq{} and \dReq{} are exactly the same as
shown in Figure \ref{msi-template}, except that whenever the state of a child
$c'$ is read, it should be replaced syntactically by $this.dir[c'][a]$.

@@

\begin{algorithmic}
\State // Upgrading $this$ by sending request to parent and getting a response
\Proc {\uReqL}{$p, a, x$}
  \If {$this$ is LLC \&\& $this.state[a] = I$}
    \State \send{} \Req{this}{Memory}{a}{M};
    \State \receive{} \Data{Memory}{this}{a}{d};
    \State $this.data[a] \gets d$;
    \State $this.state[a] \gets M$;
  \Else
  \State \send{} \Req{this}{p}{a}{x};
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndIf
\EndProc
\State // Downgrading $this$' child by sending request to child and getting a
response
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  %\While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  %\EndWhile
\EndProc
\State // $this$ finally upgrading the directory state of its child and sending
a response to its child
\Proc {\uResp}{$c, a, x$}
  \State \textbf{if} ($this.dir[c][a] = I$)
  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
  \State $this.dir[c][a] \gets x$;
\EndProc
\State // $this$ finally downgrading its state and sending a response to its
parent
\Proc {\dResp}{$p, a, x$}
  \State \textbf{if} ($isModified(this.state[a])$)
  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\caption{Distributed MSI protocol. In all the methods $p$ is the parent of
$this$, $c$ is a child of $this$, and the methods work on address $a$ to change
its state (or directory) to $x$. This function only shows the auxiliary
functions and the handler for voluntary responses from children. The handler
for requests are exactly the same as in Figure \ref{msi-template}, with every
occurrence of $c'.state[a]$ replaced by $this.dir[c'][a]$}
\label{realistic}
\end{figure}

@@

Now that we have established the vocabulary of threads, we will show the
scheduling algorithm that we generate.

\subsection{Scheduling algorithm}

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=.4]{Scheduler}
\caption{Scheduler or the cache controller}
\label{Scheduler}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

The cache controller, or, extending the thread analogy further, the scheduler
is shown in Figure \ref{Scheduler} We list the scheduling algorithm that
enables the designer to have an illusion of non-suspensive threads while
guaranteeing deadlock freedom in Figure \ref{scheduler}.

\begin{figure}
\small
%\begin{boxedminipage}{\linewidth}
\begin{alltt}
\normalfont
\lWhile{} (\(True\)) \bopen
      \lIf (\Resp{c}{n}{a}{x} is in the input response channel
             from children, where \(n = c.parent\)) \bopen
            \lIf (suspended thread \(t\) is waiting for 
                  a response from \(c\) for address \(a\)) \bopen
                   \resume{} \(t\);
            \bclose \lElsIf ($n$ has a free resource for responses) \bopen
                   \receive{} \Resp{c}{n}{a}{x};
                   \start{} \dRespL(\(c, n, a, x\));
            \bclose
      \bclose \lElsIf (\Resp{p}{n}{a}{x} is in the input channel
                       from the parent, where \(p = n.parent\)) \bopen
                   // There must be a suspended thread \(t\)
                   waiting for a response from \(p\) for address \(a\)
                   \resume{} \(t\);
            \bclose
      \bclose \lElsIf (suspended thread \(t\) is ready to execute)
             \resume{} \(t\);
      \bclose \lElsIf (\Req{c}{n}{a}{x} is in the input request
                      channel from children, where \(n = c.parent\)) \bopen
             \lIf ((\textbf{!}(address \(a\) has been chosen for eviction by 
                       some other thread in \(n\) and has not been evicted) \textbf{or}
                     (a thread in \(n\) is handling a request for address \(a\))) \textbf{and}
                    \(n\) has a free resource for requests from children) \bopen
                   \receive{} \Req{c}{n}{a}{x};
                   \start{} \uReq(\(c, n, a, x\));
             \bclose
      \bclose \lElsIf (\Req{p}{n}{a}{x} is in the input channel
                        from the parent, where \(p = n.parent\)) \bopen
             \lIf (\textbf{!}((address \(a\) has been chosen for eviction by 
                       some other thread in \(n\) and has not been evicted) \textbf{or}
                     (a thread in \(n\) is handling a request from \(p\)
                       for address \(a\)) \textbf{or}
                     (a thread \(t\) in \(n\) is handling a request from child \(c\) for
                       address \(a\) and \(t\) is not waiting for a response from \(p\))) \textbf{and}
                    \(n\) has a free resource for requests from parent) \bopen
                   \receive{} \Req{p}{n}{a}{x};
                   \start{} \dReq(\(p, n, a, x\));
             \bclose
      \bclose
\bclose
\end{alltt}
%\end{boxedminipage}
\caption{Implementation of a scheduler for cache node $n$}
\label{scheduler}
\end{figure}

It should be noted that if a request message arrives, even if there are
resources available for starting the processing of the message, the request is
sometimes not handled, \ie the scheduler does not start a new thread for
handling the message. Intuitively, this is to guarantee that only one thread is
modifying a particular address in the cache -- till the thread finishes its
work, no other thread should be allowed to modify that address. If another
thread is allowed to modify the address, then the assumptions that the first
thread has about the state of the address will be voided. Either the thread has
to restart from the beginning, or we can avoid this scenario by not scheduling
the second thread. The only exception to this rule is when a thread has been
suspended by sending a request to the parent and another request from the parent
for the same address is received. In this case, a new thread has to be created
for the request from the parent to avoid deadlocks like the ones shown in Figure
\ref{unknown}.

The scheduler must also ensure that it does some fair arbitration of threads
that are suspended. If a particular address remains suspended forever, it will
lead to starvation. The static priority that we present in the algorithm -- a
suspended thread that is ready to run always gets priority over creating new
threads for request messages, will ensure that no thread gets starved.

\begin{figure}\small
\begin{requirement}
An incoming request from the parent should not be blocked by any
incoming request from its children.\label{cReqNoBlockPReq}
\end{requirement}
\begin{requirement}
A response should not be blocked by a request.
\label{reqNoBlockResp}
\end{requirement}
\caption{Blocking restrictions}
\label{blocking}
\end{figure}

Note that we present three incoming channels to the scheduler and three incoming
channels to the scheduler. This is to ensure that a message from one of these
channels does not block a message in another queue. The exact requirements for
which messages are allowed to block are stated in Figure \ref{blocking}.

To avoid blocking, a scheduler has to have at least one dedicate resource
(MSHR) for for handling requests and responses from the parent, and one for
handling responses from children. There should at least be one more since
requests from children can not use the other two.

The channels that we depict in Figure \ref{Scheduler} are all FIFO queues. But
there is an ordering requirement even between queues between requests and
responses sent from the children. We specify the minimum ordering requirements
in Figure \ref{order}. Though the figure specifies FIFO requirements for every
address, an implementation which ensures the corresponding FIFO requirements
overall, even between addresses, will guarantee the stated requirements.

\subsection{Illustrating the need for FIFO queues and blocking restrictions}
We will illustrate the need to avoid blocking using the example of a 2-level
cache hierarchy with two L1 caches $c_1$ and $c_2$, and one shared L2 cache
$p$. Each core can have several outstanding memory requests.

Let's say both $c_1.state[a] = p.dir[c_1][a] = S$ and $c_2.state[a] =
p.dir[c_2][a] = S$. Both $c_1$ and $c_2$ get store requests from the processor
for address $a$, and they send upgrade-to-$M$ requests to $p$.  $p$ receives the
request from $c_2$ first. $p$ sends a downgrade-to-$I$ request to $c_1$.

Let's say some request from the core is blocking $c_1$ from receiving the
request. $c_1$ will never respond back to $p$. But $p$ will be waiting in
procedure \dReqL{}, and will never start handling the request from $c_1$. This
will create a deadlock.  The same deadlock is also created when $c_1$ is out of
thread resources to create a thread to handle the downgrade request from $p$.
Figure \ref{unknown} depicts this scenario.

Let's extend this scenario further. Let $c_1$ send a downgrade-to-$I$ response
to $p$. Let's assume that the downgrade-to-$I$ response from $c_1$ is being
blocked. This means $p$ doesn't receive the downgrade-to-$I$ response, leading
to the same deadlock. This is also the same if there isn't a dedicated resource
in $p$ to handle the downgrade response.

\floatstyle{boxed}
\restylefloat{figure}

@@

\subsection{Suspension due to eviction}
A thread can also get suspended if it needs to evict another line because of
lack of space in the cache for the address corresponding to the thread. This is
because, an address that is currently being processed by another thread should
not be evicted. Another point to note is that an address which is already
selected for eviction, but is waiting for the children to downgrade to $I$,
should not be selected for eviction by another address.

### 2012-11-29 paper/isca40format/Implementation.tex
before reorging scheduler etc

Since responses from the parent $p$ of a cache node $c$ should not be blocked
at the cache node by a request from $c$'s child $c'$ (Requirement
\ref{reqBlockResp2}), these responses can not use the same channel as that used
by a request from $c'$ to $c$. However, they can use the same virtual channel
used by the requests from $p$ to $c$, $ReqChannel\langle x, ToC \rangle$, where
$x$ is the level of $p$. This is because no blocking restrictions of Figure
\ref{blocking} are violated.

@@

\subsubsection{Bits required for the directory:}
\begin{wrapfigure}{r}{.4\linewidth}
\begin{tabularx}{\linewidth}{|c|X|}
\hline
State & Directory \\
\hline
$M$ & one $M$, rest $I$\\
$M$ & some $S$, rest $I$\\
$S$ & some $S$, rest $I$\\
$I$ & all $I$\\
\hline
\end{tabularx}
\caption{Possible values of coherence state and directory for each cache line}
\label{encoding}
\end{wrapfigure}
The directory of a node store its version of the state for each of the node's
children. However, one can use the fact that for a node $n$, if $n.dir[c][a] =
M$, then $\forall c' \neq c, n.dir[c'][a] = I$. This comes from Invariants
\ref{singleWriter} and \ref{conservative}. Thus, storing just the index of the
cache is sufficient if $n.dir[c][a] = M$ for some cache $c$. Figure
\ref{encoding} shows the possible values of the directory and coherence state
for each line in the cache. It can clearly be seen that this requires
$max(\lceil log_2(n)\rceil, n) + 2$ bits per line, for the most efficient
encoding.

\subsubsection{MSHRs to implement threads:}
The threads we describe essentially represent state stored in Miss Status Handling Registers (MSHRs). To
satisfy the requirements of Figure \ref{dedicated}, there must at least be 3
MSHR registers, in which one can be used only for messages from the parent, one
only for responses (both from children and the parent) and one general purpose
MSHR. We used the term threads to describe our protocol because we give the
illusion of only one thread executing in the system though the threads are
actually suspensive. Hence the software terminology fits our model quite well.

>>>>

Since responses from the parent $p$ of a cache node $c$ should not be blocked
at the cache node by a request from $c$'s child $c'$, these responses can not
use the same channel as that used by a request from $c'$ to $c$. However, they
can use the same virtual channel used by the requests from $p$ to $c$,
$ReqChannel\langle x, ToC \rangle$, where $x$ is the level of $p$. This is
because no blocking restrictions of Figure \ref{blocking} are violated.

@@

\subsubsection{Bits required for the directory:}
\floatstyle{plain}
\restylefloat{figure}
\begin{wrapfigure}{r}{.4\linewidth}
\begin{tabularx}{\linewidth}{|c|X|}
\hline
State & Directory \\
\hline
$M$ & one $M$, rest $I$\\
$M$ & some $S$, rest $I$\\
$S$ & some $S$, rest $I$\\
$I$ & all $I$\\
\hline
\end{tabularx}
\caption{Possible values of coherence state and directory for each cache line}
\label{encoding}
\end{wrapfigure}
The directory of a node store its version of the state for each of the node's
children. However, one can use the fact that for a node $n$, if $n.dir[c][a] =
M$, then $\forall c' \neq c, n.dir[c'][a] = I$. This comes from Invariants
\ref{singleWriter} and \ref{conservative}. Thus, storing just the index of the
cache is sufficient if $n.dir[c][a] = M$ for some cache $c$. Figure
\ref{encoding} shows the possible values of the directory and coherence state
for each line in the cache. It can clearly be seen that this requires
$max(\lceil log_2(n)\rceil, n) + 2$ bits per line, for the most efficient
encoding.
